Stability and Performance of Coalitions of Prosumers Through Diversification in the Smart Grid

### 2015-11-19 Appendix.tex
update

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[french, english]{babel}
\usepackage{cite}
\usepackage{caption}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{subcaption}
\usepackage{array}
\usepackage{color}
\usepackage{float}
\usepackage{multicol}
\usepackage[]{algorithm2e}
\usepackage{hyperref}
\usepackage{fancyhdr}
%
% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

\appendix
\subsection{Algorithms}

@@

\bibliographystyle{IEEEtran}  
\bibliography{Article}
\end{document}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[french, english]{babel}
\usepackage{cite}
\usepackage{caption}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{subcaption}
\usepackage{array}
\usepackage{color}
\usepackage{float}
\usepackage{multicol}
\usepackage[]{algorithm2e}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{multicol}

%
% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

\onecolumn
\centering
\appendix
   \subsection{Notations}

	\begin{tabular}{ | l | r | }
   						N & Number of Prosumers \\
   						$ P_i^P(t) $ & Available production of agent i at time t \\
   						$ P_i^D(t) $ & Consumption of agent i at time t \\ 
  						$ P_i(t) = P_i^P(t) - P_i^D(t) $ & Available extra-production of agent i at time t \\
  						$  WT_i $ & Wind turbine set of agent i \\
  						$ PV_i $ & Solar array set of agent i \\
  						$ DER_i = WT_i \cup PV_i $ & DER of agent i \\
  						$ \nu_i(t) $ & wind speed (in $ m s^{-2} $) at agent i's location and time t \\
  						$ \Psi_i(t) $ & solar irradiance (in $ W m^{-2} $) at agent i's location and time t \\
  						$ \mathcal{F}_{WT} $ & power curve for the wind turbines \\
  						$ \mathcal{F}_{PV} $ & power curve for the solar arrays \\
   						$ P_S^{CRCT} $ & Contract value of coalition S \\
   						$ \phi \in \left[ 0,1 \right] $ & Reliability Threshold \\
   						$ P^{MIN} $ & Minimum Contract value \\
   						$ \alpha $ & Parameter that controls the coalitions sizes \\
   						$ \epsilon $ & Correlation graph filtering threshold \\
   						$ N_{COAL} $ & Number of coalitions \\
   						$ \mathcal{R}_S $ & Resilience of coalition S \\
   						$ \lambda $ & unitary price rate for electricity \\
   						$ \delta_S(i) $ & marginal utility of agent i in coalition S \\
   						$ \rho_{ij} $ & Pearson's correlation coefficient between $ P_i$ and $P_j$ \\
   						$ \Theta_k(G) $ & set of non overlapping cliques of size k in graph G \\
   						$ \mathcal{R}_S $ & resilience of coalition S
 	\end{tabular}

\twocolumn
\subsection{Algorithms}

@@

\bibliographystyle{IEEEtran}  
\bibliography{Article}

\end{document}

### 2015-11-19 Article.tex
update

The recent developments in the smart grid area, such as demand side management mechanisms, storage efficiency, or communication architectures and algorithms to name a few, sketch a new power system. The rigid top-down centralized architecture of today’s power grids is expected to evolve in more distributed and flexible systems. Bidirectional power flows, storage, and efficient use of information are some of the key principles that will enable this transition. Since production could be located down to the very end of the distribution networks, nodes that were simply pure loads yesterday could behave tomorrow as generators or loads, further complicating energy usages [5]. 
In this paper, we focus on these "nodes" in the distribution network that can produce and consume electricity. More precisely, we consider a set of agents that own both renewable distributed energy resources (DER) and electrical loads. The production of an agent can be, of course, used to meet its own demand, but in cases when it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "prosumer" [6]. A key point of this work is to merge the interests of the grid operator with the prosumers in a single utility function. While the latter intend to maximize their benefits with higher production contracts, the operator primary concerns are related to the quality and stability of the injections. The coalitions that we wish to form should thus be both stable and productive. This compromise can be difficult to obtain in real situations, especially when renewable generation and complex spatiotemporal correlation happen.

More precisely, in our model, any entity that wishes to sell its production on some energy market has to estimate and announce a production capacity for an upcoming period of time. If a contract is bought, the entity commits to injecting, at any time of the contracted period, the announced amount of power, under financial penalties if it fails. Since prosumers use exclusively renewables, contracts come naturally with some amount of risk due to the intermittent nature of most renewables. Using storage as buffers as to reduce risks is a popular approach. Without storage and proper control, the over-producing state of the agents, and thus the power they inject in the grid, might be rather unstable. This is unacceptable for the grid operator that cannot ensure system stability if it has to deal with numerous small unpredictable entities.

Forming coalitions/aggregations is an envisaged solution to both the number of entities soliciting the operator, and their high variability. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performance on the communication side \cite{Negeri2012}. On the other hand, it is known that diversification of the assets is a way of minimizing risks when building a portfolio. One thus expect a more stable and predictable energy production for a coalition than for a single agent. Nevertheless, all coalitions are not equally stable or productive. Special attention should thus be paid to the aggregation step. Recall that prosumers have both a consumption and a production component, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

@@

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity prices for instance), and take actions accordingly (demand side management programs). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are nonetheless being conducted and data are progressively made public \cite{ISSDA}. In this paper, we use weather quantities like wind speed or solar radiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following (see block 1 of fig. \ref{fig:process}). A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. More formally, we denote by $ P_{i}(t) $ the instantaneous extra-production of agent i at time t :

@@

Note that a prosumer i is defined by his zone $ Z_{i} $ as well as the sets $ DER_{i} $ and $ load_{i} $. That is, a prosumer can be configured to represent anything from a single wind-turbine for instance ($ DER_{i} = \{ WT_{0} \} $ and $ load_{i} = \emptyset $) to a pure load ($ DER_{i} = \emptyset $ and $ load_{i} = \{ L_{0} \} $) through more complex combinations. In practice, we use random configurations for the agents. In the rest of the paper, we use french weather data \cite{Infoclimat} starting in January 2006 and ending in December 2012, with a sampling frequency of three hours, and generate N timeseries of extra-production over this date range.

@@

This section provides most of the notations and introduces important concepts for the rest of the paper. We consider a set $ \mathcal{A} = \{a_{1},a_{2},...,a_{N} \} $ of N prosumers of the distribution network. Each agent is configured randomly and we simulate its extra-production $ P_{i}(t),\ \forall i \in \mathcal{A} $ from 2006 to 2012. Based on these historical values, our objective is now to form groups of prosumers so that the global power production resulting from the superposition of individual's extra-productions be both sufficiently high and predictable. Let $ P_{S}(t) = \sum_{i \in S} P_{i}(t) $ be the extra-production of coalition S at time t.

>>>>

The recent developments in the smart grid area, such as demand side management mechanisms, storage efficiency, or communication architectures to name a few, sketch a new power system. The rigid top-down centralized architecture of today’s power grids is expected to evolve in more distributed and flexible systems. Bidirectional power flows, storage, and efficient use of information are some of the key principles that will enable this transition. Since production could be located down to the very end of the distribution networks, nodes that were simply pure loads yesterday could behave tomorrow as generators or loads, further complicating energy usages [5]. 
In this paper, we focus on these "nodes" in the distribution network that can produce and consume electricity. More precisely, we consider agents owning both renewable distributed energy resources (DER) and electrical loads. The production of an agent can be used to meet its own demand, but in cases when it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "prosumer" [6]. A key point of this work is to merge the interests of the grid operator with the prosumers in a single utility function. While the latter intend to maximize their benefits with better contracts, the operator concerns are related to the quality and reliability of the injections.

%The coalitions that we wish to form should thus be both stable and productive. This compromise can be difficult to obtain in real situations, especially when renewable generation and complex spatiotemporal correlation happen.

%More precisely, in our model, any entity that wishes to sell its production on some energy market has to estimate and announce a production capacity for an upcoming period of time. If a contract is bought, the entity commits to injecting, at any time of the contracted period, the announced amount of power, under financial penalties if it fails. Since prosumers use exclusively renewables, contracts come naturally with some amount of risk due to the intermittent nature of most renewables. Using storage as buffers as to reduce risks is a popular approach. Without storage and proper control, the over-producing state of the agents, and thus the power they inject in the grid, might be rather unstable. This is unacceptable for the grid operator that cannot ensure system stability if it has to deal with numerous small unpredictable entities.

We rely on a market representation of the interactions. Coalitions of prosumers estimate a production capacity for an upcoming period and announce it on the market. If some other entity buys the contract, the coalition has to inject exactly the contracted amount of power at any time. Obviously, deviations will cause financial penalties to the coalitions, such that respecting the contracts is paramount. Since prosumers use exclusively renewables, we are particularly concerned with the probability that the available capacity of a coalition falls below its contract value. In this case, storage or demand side managment might be solutions to avoid penalties, but this probability should still be maintained low.

We focus on coalitions because it appears as a solution to both the number of prosumers communicating with the operator as well as their high variabilities. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performance on the communication side \cite{Negeri2012}. On the other hand, it is known that diversification of the assets is a way of minimizing risks when building a portfolio. We thus expect more stable and predictable energy productions for coalitions than for single agents. Nevertheless, all coalitions are not equally stable or productive. Special attention should thus be paid to the aggregation step. Recall that prosumers have both consumption and production components, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

@@

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity prices for instance), and take actions accordingly (demand side management programs). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are nonetheless being conducted and data are progressively made public \cite{ISSDA}. In this paper, we use weather quantities like wind speed or solar radiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following (see block 1 of fig. \ref{fig:process}). A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. More formally, we denote by $ P_{i}(t) $ the instantaneous available extra-production of agent i at time t :

@@

Note that a prosumer i is defined by his zone $ Z_{i} $ as well as the sets $ DER_{i} $ and $ load_{i} $. That is, a prosumer can be configured to represent anything from a single wind-turbine for instance ($ DER_{i} = \{ WT_{0} \} $ and $ load_{i} = \emptyset $) to a pure load ($ DER_{i} = \emptyset $ and $ load_{i} = \{ L_{0} \} $) through more complex combinations. In practice, we use random configurations for the agents. In the rest of the paper, we use french weather data \cite{Infoclimat} starting in January 2006 and ending in December 2012, with a sampling frequency of three hours, and generate N timeseries of extra-production over this date range. Note that this low sampling frequency might hide short time intermittencies (see \cite{Anvari2015} for more information). Studying the evolution of the correlation structure with the sampling frequency is out of the scope of this paper, but might lead to interresting results.

@@

This section provides most of the notations and introduces important concepts for the rest of the paper. We consider a set $ \mathcal{A} = \{a_{1},a_{2},...,a_{N} \} $ of N prosumers of the distribution network. Each agent is configured randomly and we simulate its extra-production $ P_{i}(t),\ \forall i \in \mathcal{A} $ from 2006 to 2012. Based on these historical values, our objective is now to form groups of prosumers so that the global power production resulting from the superposition of individual's extra-productions be both sufficiently high and predictable. Let $ P_{S}(t) = \sum_{i \in S} P_{i}(t) $ be the available extra-production of coalition S at time t.

### 2015-11-19 letter.tex
update

@@

\textbf{[Answer]} In this paper, we consider the case of a large penetration of renewables in the distribution network. A significant portion of the production is assumed to come from these coalitions of prosumers, in such a way that the grid operator is scheduling the plants according to the contract values. If coalitions do not satisfy their contracts by under-producing, it might be the case that the grid operator has to re-balance consumption and production by turning on fast ramping generators. This might generates high operational costs for the grid operator.

>>>>

We added the following text at the end of section 3 :

\textit{Note that this low sampling frequency might hide short time intermittencies (see \cite{Anvari2015} for more information). Studying the evolution of the correlation structure with the sampling frequency is out of the scope of this paper, but might lead to interresting results.}

@@

\textbf{[Answer]} We agree with the reviewer that, in the case of a large penetration of renewables in the distribution network, the production capacity should be larger than for traditional power plants. This is the consequence of the uncertainty that comes with most renewables. It results from this point that, at certain periods, the production might be drastically larger than the consumption, which might indeed cause various problems in the network. Storage, demand-side managment, inter-connection with neighboring countries, as well as energy markets are often assumed to provide ways of stabilizing the grid in such situations.

Nevertheless, we want to point out that, in this paper, we focus on the market side of the system rather than the grid pure stability. More precisely, we consider the formation of coalitions of prosumers that seek to sell energy to some entity (grid operator, aggregators). If this entity buys a contract, the coalition commits to injecting exactly $ P_S^{CRCT} $ :

\begin{equation}
\forall t,\ P_S^{injected}(t) = P_S^{CRCT} 
\end{equation}

Obviously, if the actual injected power deviates from the contract, the coalition should pay penalties proportional to $ \sum_t \left| P_S^{injected}(t) - P_S^{CRCT} \right| $. However, within this framework, we build the coalitions by looking at their available power $P_S(t)$, i.e, the maximum production that a coalition can inject at time t : $ P_S^{injected}(t) \leq P_S(t) $. 
In this situation, we believe that the risk $ Pr[P_S(t) < P^{CRCT} ] $ is a key quantity since, in this case, the contract cannot be respected. In cases where the available production is higher than the contract $ P_S(t) > P^{CRCT} $, it might be possible to shut some renewables down, or to use storage, to maintain the power injection at the contract value.

We added the following text in the introduction as to clarify this point :

\textit{We rely on a market representation of the interactions. Coalitions of prosumers estimate a production capacity for an upcoming period and announce it on the market. If some other entity buys the contract, the coalition has to inject exactly the contracted amount of power at any time. Obviously, deviations will cause financial penalties to the coalitions, such that respecting the contracts is paramount. Since prosumers use exclusively renewables, we are particularly concerned with the probability that the available capacity of a coalition falls below its contract value. In this case, storage or demand side managment might be solutions to avoid penalties, but this probability should still be maintained low.}
